A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-11-05 latex/main.tex
Updates from Overleaf

Millions of SARS-CoV-2 genome sequences were collected 
during the COVID-19 pandemic, forming a dataset of unprecedented richness.
Estimated genealogies are fundamental to
understanding this ocean of data, 
and form the primary input to many downstream analyses.
A basic assumption of methods to infer genealogies from viral 
genetic data is that recombination is negligible and 
the genealogy is a tree.
However, recombinant lineages have risen to global prevalence 
and simple tree representations are therefore 
incomplete and potentially misleading.
We present sc2ts, a method to infer reticulate genealogies 
as an Ancestral Recombination Graph (ARG) in real time at pandemic scale.
We infer an ARG for 2.48 million SARS-CoV-2 genomes,
and show that this deeply validated resource clarifies
the relationships between recombinant lineages,
quantifies the rate of recombination over time
and provides a conservative
lower bound on detectable recombination.
This ARG and rich supporting software ecosytem provide
a strong platform for future research.

@@

Here we present sc2ts, a new method to infer ARGs for SARS-CoV-2  
that incrementally adds samples over time in a similar manner 
to UShER, but also automatically detects and incorporates recombination
in real-time. Based on the Viridian dataset, a freely available  
resource purged of various systematic errors and artifacts~\cite{Hunt2024},
we infer an ARG with 2.48 million whole genome sequences.
We show that this ARG is highly congruent with the UShER phylogeny built 
from the same dataset as well as the Pango nomenclature system~\cite{Rambaut2020},
and also accurately captures evolutionary features such as mutational 
spectra~\cite{Bloom2023} and deletions\cite{Li2023indels}.
We show that sc2ts automatically and accurately detects details about
known recombinant lineages that were painstakingly collated in a community 
process, and provides novel insights into the relationships between these
recombinant lineages. We compare the recombination events detected by 
sc2ts with the state-of-the-art combination of UShER and 
RIPPLES~\cite{Turakhia2022}, showing that sc2ts is more sensitive,
detecting more of the well-characterized recombination events 
curated by the Pango lineage designation community.
The output of sc2ts---incorporating
genealogical relationships, point mutations, deletions, and 
recombination---provides a basis for the study of SARS-CoV-2
that systematically accounts for all of these evolutionary forces.
This ARG, coupled with the rich suite of supporting software
in the tskit~\cite{Ralph2020} and VCF-Zarr~\cite{Czech2025} ecosystems,
enables pandemic-scale analysis on a laptop. 
These resources provide a strong platform for future research,
helping us to learn from the vast troves of data collected during the
pandemic, and to prepare for the next.

>>>>

Millions of SARS-CoV-2 genome sequences were collected 
during the COVID-19 pandemic, forming a dataset of unprecedented richness.
Estimated genealogies are fundamental to
understanding this ocean of data, 
and form the primary input to many downstream analyses.
A basic assumption of methods to infer genealogies from viral 
genetic data is that recombination is negligible and 
the genealogy is a tree.
However, recombinant lineages have risen to global prevalence 
and simple tree representations are therefore 
incomplete and potentially misleading.
We present sc2ts, a method to infer reticulate genealogies 
as an Ancestral Recombination Graph (ARG) in real time at pandemic scale.
We infer an ARG for 2.48 million SARS-CoV-2 genomes,
and show that this deeply validated resource clarifies
the relationships between recombinant lineages,
quantifies the rate of recombination over time
and provides a conservative
lower bound on detectable recombination.
This ARG and rich supporting software ecosystem provide
a strong platform for future research.

@@

Here we present sc2ts, a new method to infer ARGs for SARS-CoV-2  
that incrementally adds samples over time in a similar manner 
to UShER, but also automatically detects and incorporates recombination
in real-time. Based on the Viridian dataset, a freely available  
resource purged of various systematic errors and artifacts~\cite{Hunt2024},
we infer an ARG with 2.48 million whole genome sequences.
We show that the non-recombinant portions of this ARG are highly congruent with the UShER phylogeny built 
from the same dataset as well as the Pango nomenclature system~\cite{Rambaut2020},
and that the ARG accurately captures evolutionary features such as mutational 
spectra~\cite{Bloom2023} and deletions\cite{Li2023indels}.
We show that sc2ts automatically and accurately detects details about
known recombinant lineages that were painstakingly collated in a community 
process, and provides novel insights into the relationships between these
recombinant lineages. We compare the recombination events detected by 
sc2ts with the state-of-the-art combination of UShER and 
RIPPLES~\cite{Turakhia2022}, showing that sc2ts is more sensitive,
detecting more of the well-characterized recombination events 
curated by the Pango lineage designation community.
The output of sc2ts---incorporating
genealogical relationships, point mutations, deletions, and 
recombination---provides a basis for the study of SARS-CoV-2
that systematically accounts for all of these evolutionary forces.
This ARG, coupled with the rich suite of supporting software
in the tskit~\cite{Ralph2020} and VCF-Zarr~\cite{Czech2025} ecosystems,
enables pandemic-scale analysis on a laptop. 
These resources provide a strong platform for future research,
helping us to learn from the vast troves of data collected during the
pandemic, and to prepare for the next.
